---
title: "Checking whether `skinny` is suitable as a projection pursuit index"
format:
  html:
    code-fold: true
date: 2026-06-23
execute:
  warning: false
  message: false
  cache: true
license: "All Rights Reserved"
copyright: "Copyright © 2026 Tina Rashid Jafari. All rights reserved."
---

## Goal

Before using `skinny` inside a guided tour, I first check whether it behaves like a useful projection pursuit index.

For data matrix $X \in \mathbb{R}^{n \times p}$ and projection basis $A \in \mathbb{R}^{p \times 2}$, the projected data are

$$
Y = XA.
$$

The projection pursuit index is then calculated on $Y$.


Here I compare two versions of the index:

- `skinny(rescale = FALSE)`: the original skinny index.
- `skinny(rescale = TRUE)`: a rescaled version designed to reduce the influence of values that commonly arise from Gaussian noise.

In previous work, I estimated the distribution of the skinny index under bivariate Gaussian noise across different sample sizes. Using the estimated upper tail of this noise distribution, I derived a sample-size-dependent threshold

$$
\ell(n)=0.72-0.19\log(n)+0.02\{\log(n)\}^2,
$$

The rescaled index is then defined as

$$
I_{\text{rescaled}}
=
\max \left(
0,
\frac{I - \ell(n)}
     {1 - \ell(n)}
\right),
$$

where $I$ is the original stringy05 value.

This transformation preserves the upper end of the index while shrinking values that are likely to arise from Gaussian noise. Values below the estimated noise threshold are mapped to zero, and values above the threshold are linearly rescaled to the interval $[0,1]$.

The motivation is that random noise projections can sometimes produce non-zero stringy values, potentially making optimisation more difficult. By incorporating an estimated noise threshold, the rescaled version aims to increase separation between structured projections and noise projections, allowing the optimiser to focus more strongly on projections containing genuine skinny structure.


```{r}
# Load packages
library(tourr)
library(cassowaryr)
library(spinebil)
library(ferrn)
library(dplyr)
library(tidyr)
library(purrr)
library(ggplot2)
```

## Define raw and rescaled index functions


```r
rescale_skinny <- function(z, n) {
  lb <- 0.72 - 0.19 * log(n) + 0.02 * (log(n)^2)
  pmax(0, (z - lb) / (1 - lb))
}

skinny_index <- function(rescale = FALSE) {
  function(mat) {

    z <- cassowaryr::sc_skinny(mat[, 1], mat[, 2])

    if (rescale) {
      z <- rescale_skinny(z, nrow(mat))
    }

    z
  }
}

skinny_index_raw <- skinny_index(rescale = FALSE)
skinny_index_rescaled <- skinny_index(rescale = TRUE)
```


## Generate datasets

I use four datasets:

1. a hidden polynomial dataset, with the signal in variables 2 and 3;
2. a pipe dataset where the structure is in the final two variables;
3. a sine dataset;
4. a spiral dataset.

### Polynomial data

```{r}
make_poly_data <- function(n = 300, p = 4, degree = 2, seed = 1050,
                           signal_noise_sd = 0.005) {
  set.seed(seed)

  t <- seq(-1, 1, length.out = n)
  signal <- poly(t, degree = degree, raw = TRUE)

  x <- matrix(rnorm(n * p), nrow = n, ncol = p)

  x[, 2] <- signal[, 1] + rnorm(n, sd = signal_noise_sd)
  x[, 3] <- signal[, 2] + rnorm(n, sd = signal_noise_sd)

  colnames(x) <- paste0("V", seq_len(p))

  as.data.frame(x)
}

dat4 <- scale(make_poly_data(n = 300, p = 4, degree = 2, seed = 1050))
```

The polynomial signal is in variables 2 and 3.

```{r}
basis_poly_true <- spinebil::basis_matrix(2, 3, 4)
basis_poly_noise <- spinebil::basis_matrix(1, 4, 4)
```

### Pipe, sine, and spiral data

For these `spinebil` datasets, I use six variables. The structured variables are `V5` and `V6`, and the nuisance variables are `V1`, `V2`, `V3`, and `V4`.

```{r}
set.seed(1050)

pipe_dat <- spinebil::pipe_data(300, 6) |> scale()
sine_dat <- spinebil::sin_data(300, 6, 1) |> scale()
spiral_dat <- spinebil::spiral_data(300, 6) |> scale()

basis_spinebil_true <- spinebil::basis_matrix(5, 6, 6)
basis_spinebil_noise1 <- spinebil::basis_matrix(1, 2, 6)
basis_spinebil_noise2 <- spinebil::basis_matrix(3, 4, 6)
```



## Plot the true structures

Before using the guided tour, I first plotted the signal variables directly. This is the target structures that I want the guided tour to find.

```{r}
poly_plot <- as.data.frame(dat4) |>
  transmute(x = V2, y = V3, data = "polynomial")

pipe_plot <- as.data.frame(pipe_dat) |>
  transmute(x = V5, y = V6, data = "pipe")

sine_plot <- as.data.frame(sine_dat) |>
  transmute(x = V5, y = V6, data = "sine")

spiral_plot <- as.data.frame(spiral_dat) |>
  transmute(x = V5, y = V6, data = "spiral")

bind_rows(poly_plot, pipe_plot, sine_plot, spiral_plot) |>
  ggplot(aes(x = x, y = y)) +
  geom_point(size = 0.8, alpha = 0.7) +
  facet_wrap(~ data, scales = "free") +
  theme(
    aspect.ratio = 1,
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "True structured projections"
  )
```


## Direct check: signal versus noise


```{r}
rescale_skinny <- function(z, n) {
  lb <- 0.72 - 0.19 * log(n) + 0.02 * (log(n)^2)
  pmax(0, (z - lb) / (1 - lb))
}

skinny_index <- function(rescale = FALSE) {
  function(mat) {

    z <- cassowaryr::sc_skinny(mat[, 1], mat[, 2])

    if (rescale) {
      z <- rescale_skinny(z, nrow(mat))
    }

    z
  }
}

skinny_index_raw <- skinny_index(rescale = FALSE)
skinny_index_rescaled <- skinny_index(rescale = TRUE)


direct_check <- tibble(
  data = c("polynomial", "pipe", "sine", "spiral"),
  raw_signal = c(
    skinny_index_raw(dat4 %*% basis_poly_true),
    skinny_index_raw(pipe_dat %*% basis_spinebil_true),
    skinny_index_raw(sine_dat %*% basis_spinebil_true),
    skinny_index_raw(spiral_dat %*% basis_spinebil_true)
  ),
  raw_noise = c(
    skinny_index_raw(dat4 %*% basis_poly_noise),
    skinny_index_raw(pipe_dat %*% basis_spinebil_noise1),
    skinny_index_raw(sine_dat %*% basis_spinebil_noise1),
    skinny_index_raw(spiral_dat %*% basis_spinebil_noise1)
  ),
  rescaled_signal = c(
    skinny_index_rescaled(dat4 %*% basis_poly_true),
    skinny_index_rescaled(pipe_dat %*% basis_spinebil_true),
    skinny_index_rescaled(sine_dat %*% basis_spinebil_true),
    skinny_index_rescaled(spiral_dat %*% basis_spinebil_true)
  ),
  rescaled_noise = c(
    skinny_index_rescaled(dat4 %*% basis_poly_noise),
    skinny_index_rescaled(pipe_dat %*% basis_spinebil_noise1),
    skinny_index_rescaled(sine_dat %*% basis_spinebil_noise1),
    skinny_index_rescaled(spiral_dat %*% basis_spinebil_noise1)
  )
)

knitr::kable(direct_check, digits = 3)
```

```{r}

plot_proj <- function(dat, basis, data, projection) {
  as.data.frame(as.matrix(dat) %*% basis) |>
    setNames(c("x", "y")) |>
    mutate(data = data, projection = projection)
}

projection_plot_data <- bind_rows(
  plot_proj(dat4, basis_poly_true, "polynomial", "signal"),
  plot_proj(dat4, basis_poly_noise, "polynomial", "noise"),
  plot_proj(pipe_dat, basis_spinebil_true, "pipe", "signal"),
  plot_proj(pipe_dat, basis_spinebil_noise1, "pipe", "noise"),
  plot_proj(sine_dat, basis_spinebil_true, "sine", "signal"),
  plot_proj(sine_dat, basis_spinebil_noise1, "sine", "noise"),
  plot_proj(spiral_dat, basis_spinebil_true, "spiral", "signal"),
  plot_proj(spiral_dat, basis_spinebil_noise1, "spiral", "noise")
)

ggplot(projection_plot_data, aes(x = x, y = y)) +
  geom_point(size = 0.8, alpha = 0.7) +
  facet_grid(projection ~ data, scales = "free")  +
  theme(
    aspect.ratio = 1,
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "Signal and noise projections"
  )
```

A useful projection pursuit index should give larger values to the signal projection than to the noise projection. The rescaled version should reduce the values assigned to noise projections.


## Rotation invariance


For a 2D guided tour, the index should measure the plane, not the orientation inside the plane. This means rotating the projected 2D data should not substantially change the index value.

```{r}

index_list <- list(
  skinny_index_raw,
  skinny_index_rescaled
)

index_labels <- c(
  "skinny raw",
  "skinny rescaled"
)

rotation_poly <- spinebil::profile_rotation(
  d = dat4 %*% basis_poly_true,
  index_list = index_list,
  index_labels = index_labels,
  n = 200
)

spinebil::plot_rotation(rotation_poly) +
  ggtitle("Rotation invariance: polynomial")
```

For the other datasets:

```{r}
rotation_pipe <- spinebil::profile_rotation(
  d = pipe_dat %*% basis_spinebil_true,
  index_list = index_list,
  index_labels = index_labels,
  n = 200
)

rotation_sine <- spinebil::profile_rotation(
  d = sine_dat %*% basis_spinebil_true,
  index_list = index_list,
  index_labels = index_labels,
  n = 200
)

rotation_spiral <- spinebil::profile_rotation(
  d = spiral_dat %*% basis_spinebil_true,
  index_list = index_list,
  index_labels = index_labels,
  n = 200
)

spinebil::plot_rotation(rotation_pipe) +
  ggtitle("Rotation invariance: pipe")

spinebil::plot_rotation(rotation_sine)  +
  ggtitle("Rotation invariance: sine")

spinebil::plot_rotation(rotation_spiral) +
  ggtitle("Rotation invariance: spiral")
```


